---
title: "Visualization in R Exercise 1: ggplot2"
author: "Andrea Batch"
date: "August 10, 2020"
output: html_document
---

```{r setup, include=FALSE}

#You must change this!
WORKING_DIRECTORY <- 'CHANGE THIS TO YOUR WORKING DIRECTORY (top level folder where you cloned git repo)'

#Set working directory (also do this for knit)
setwd(WORKING_DIRECTORY)
knitr::opts_knit$set(root.dir = WORKING_DIRECTORY)

knitr::opts_chunk$set(echo = TRUE)

.libPaths('\\\\serv03cl\\data\\APPS\\R\\libraries_3_6_3_C')
#Load the ggplot2 package
library(ggplot2)
```

## Try out the example from the slides
```{r slide_demo}
p_iris <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, colour = Species)) + 
	geom_point() + 
	stat_smooth(method = lm) + 
	scale_x_log10() + 
	scale_y_log10() 

p_iris

```

## Acquire dataset

We want to start by loading our dataset. We will use the 2nd Quarter 2020 (Advance Estimate) and Annual Update release as downloaded from the BEA site.

If we look at the data, we notice that we will need to skip some rows--4 rows--at the beginning of the file, or else we will get an error, so we use the "skip" option.

We also observe that loading it with headers creates variable names out of the years, and appends name with number as "[year].[#]" if there is more than one column for the year. We can use this to rename our variables.

```{r gdp}
data <- read.csv('data/national_gdp.csv', skip=4)
#Now look at the data
data

# Use the firstrow to rename the columns
new_names_year <- unlist( #unlist unravels list structure with depth 1
  lapply( #lapply applies a function to a list
    strsplit(names(data), split='.', fixed=T), #strsplit will return a list of vectors split from input vector on "." 
    function(x){ #this is the function being applied to strsplit return list by lapply
      return(x[1]) #return the first item ofthe vector--the year, preceded by "X" (or "Line"/"X" for first 2 columns)
  })
)

#Join new names with the first row 
new_names_qtrs <- paste0(new_names_year, as.character(unlist(data[1,])))

names(data) <- new_names_qtrs

#delete the first row!
gdp_raw <- data[-c(1), ]
gdp_raw
```

## Just for fun, let's also get quarterly population estimates.
```{r popdata}
pop <- read.csv('data/national_population.csv')

pop$Period <- paste0(substr(pop$DATE, 1, 4), 'Q', round(as.numeric(substr(pop$DATE, 6, 7))/3)+1)

pop

```

## Reshape the data for visualization

We have already made one parsing decision that has reshaped our data by omission with our skip option.

Now we are going to have to change it into a "clean" format: Every column is a variable name, every row is an observation.

We can do this more easily with the "reshape2" library.

```{r reshape}
library(reshape2)
# Melt to long format with line number and component as ID variables
gdp_long <- melt(gdp_raw, id.vars=c('LineLine', 'X'))

# Now is a good time to rename our variables
names(gdp_long) <- c('Line', 'Component', 'Period', 'Dollars')

# Remove the "X" from the period
gdp_long$Period <- gsub('X', '', gdp_long$Period, fixed = T)

# For now, we really only want top line GDP--line 1.
# We make a new data.frame from the old, ensuring our data types are correct.

# Also, we can add our pop data in now; I have been kind and set it up so that the dates are the same, but usually you would want to use a join to make sure the dates match for each row.

gdp <- data.frame(
  Period = as.character(gdp_long[gdp_long$Line==1,]$Period), 
  Dollars = as.numeric(gdp_long[gdp_long$Line==1,]$Dollars),
  Population = pop$Population
)

gdp

```

## Creating Plots

Now that we have reshaped our data, let's look at the base R "plot" function visualization:

```{r baseplot}
plot(gdp$Period, gdp$Dollars)
```

## Creating Plots and Layers with ggplot2

The base plot function returns something that's not so pretty, right?

Let's take a look at the steps involved in plotting with ggplot. Start by adding the base layer--the data and variable mappings you will use to represent your data. 

We use "group=1" here because there is no grouping variable--no nominal labels to use as "color" scale mapping (see  example demo from slides). 


```{r ggplot_base}
#Create the base layer
p_0 <- ggplot(gdp, aes(x=Period, y=Dollars, group=1))

#At first, this is just an empty canvas with axis labels and grid lines.
p_0
```

### Using ggplot layers
We need to add our visual marks (and marks' channel specifications), and graph themes and settings, layer by layer. 

Let's actually make use of our population data--set the size of the line to Dollars/Population, and then plot regular old dollars on top of that.

We immediately see that, right now, it's actually worse than base R autoplot, because we have not used any layers to modify our view.

```{r ggplot_mark}
p_1 <- p_0 + 
  geom_line(aes(size = Dollars/Population), color = "#FC4E07") +
  geom_line(size = 1.05)

p_1

```

### statistical transformation
We used stat smooth in our sepal example; let's do that again here, but instead of using a linear model, we'll use LOESS (locally estimated scatterplot smoothing).

```{r stat_layer}
p_2 <- p_1 +
  stat_smooth(method = "loess")

p_2
```



### Position scale layers
Let's start by formatting our axes using scales. Because we have so many periods, suppose we want to label a quarter every 5 years instead of every quarter:

```{r axis_scale_layer}

dates_to_show <- gdp$Period[seq(1, length(gdp$Period), by = 20)]
print(dates_to_show)

p_3 <- p_2 + 
  scale_y_continuous(labels=scales::dollar_format()) +  # Use the "dollar_format()" function from scales (ggplot dep'cy)
  scale_x_discrete(breaks = dates_to_show) # Use the dates_to_show var we defined above to skip 20 quarters

p_3

```

### Themes

These axis labels still aren't great; we can use a theme layer to rotate them. While we're at it, we may as well also make some tweaks to our background and add a chart title. 

```{r themes}
p_4 <- p_3 +
  theme_minimal() +                                      #I personally am a fan of the minimal view
  labs(title = "Nominal Gross Domestic Product",         #Add chart title, subtitle
       subtitle = "Total [Billions], Per Capita, and Projection") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1), #Rotate axis labels and shift to the right
        plot.title = element_text(hjust = 0.5),          #Center chart title
        plot.subtitle = element_text(hjust = 0.5))       #Center chart subtitle
p_4
```

### Annotation

You can add text and other visual marks freely using ggplot's layered grammar.

```{r text_annotate}
# a data frame with all the annotation info
annotation <- data.frame(
   Period = c("2007Q4","2009Q2", "2020Q2"),
   Dollars = c(20000,10000, 25000),
   label = c("Start of G.R.", "End of G.R.", "COVID Recession")
)

#Text annotation can be done several ways, including annotate(), geom_label, and geom_text()
# Here, we have chosen to use geom_text() and an annotation data.frame corresponding to GDP vars.
p_5 <- p_4 + 
  geom_text(data=annotation, aes( x=Period, y=Dollars, label=label), 
            color="black", size=3 , angle=45, 
            fontface="bold", hjust = 1)

p_5
```

We have added text, but we can also add geom annotations using annotate().
```{r shape_annotation}
p_6 <- p_5 + 
  annotate("rect", xmin = '2007Q4', xmax = '2009Q2', ymin = 0, ymax = 25000, color = "orange", fill = "orange", size = 0.5, alpha=0.4)

p_6
```

We can use the rect approach above, and we can also use a pointrange.
```{r shape_annotation_2}
p_7 <- p_6 + 
  annotate("pointrange", x = '2020Q2', y = 20000, xmin = '2020Q1', xmax = '2020Q2', ymin = 0, ymax = 25000, color = "red", size = 1.5, alpha=0.4)

p_7
```


The problem with the above--although it may be hard to tell--is that the shape annotations render on top of the other marks.
We would usually want those in the background!

So let's just bring everything together in one call.
```{r finale}
p <- ggplot(gdp, aes(x=Period, y=Dollars, group=1)) + # our data and mapping layer
  annotate("rect", xmin = '2007Q4', xmax = '2009Q2', ymin = 0, ymax = 25000, color = "orange", fill = "orange", size = 0.5, alpha=0.4) + # Our annotation layers go first!
  annotate("pointrange", x = '2020Q2', y = 20000, xmin = '2020Q1', xmax = '2020Q2', ymin = 0, ymax = 25000, color = "red", size = 1.5, alpha=.75) +
  geom_line(aes(size = Dollars/Population), color = "#FC4E07", alpha = 0.5) +  # Our two geom line (mark) layers
  geom_line(size = 1.05) +
  stat_smooth(method = "loess") + # Our stats layer 
  scale_y_continuous(labels=scales::dollar_format()) +  # Use the "dollar_format()" function from scales (ggplot dep'cy)
  scale_x_discrete(breaks = dates_to_show) + # Use the dates_to_show var we defined above to skip 20 quarters
  geom_text(data=annotation, aes( x=Period, y=Dollars, label=label),   #Add our text annotations AFTER all other
            color="black", size=3 , angle=45,                          #  geoms so it is on top of all other marks
            fontface="bold", hjust = 1) + 
  theme_minimal() +                                      #Wrap up with our themes. Minimal view theme must come first!
  labs(title = "Nominal Gross Domestic Product",         #Add chart title, subtitle
       subtitle = "Total [Billions], Per Capita, and Projection") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1), #Rotate axis labels and shift to the right
        plot.title = element_text(hjust = 0.5),          #Center chart title
        plot.subtitle = element_text(hjust = 0.5))       #Center chart subtitle

p


```