---
title: 'Visualization in R Exercise 2: ggvis'
author: "Andrea Batch"
date: "August 18, 2020"
output: html_document
---

```{r workdir, include=FALSE}

#You must change this!
WORKING_DIRECTORY <- 'CHANGE THIS TO YOUR WORKING DIRECTORY (top level folder where you cloned git repo)'
#Set working directory (also do this for knit)
setwd(WORKING_DIRECTORY)
knitr::opts_knit$set(root.dir = WORKING_DIRECTORY)
knitr::opts_chunk$set(echo = TRUE)

```

Setup some requirements
```{r setup, include=FALSE}
temp_lib_path <- paste0('C:/Users/', Sys.info()[[7]], '/Documents/temp_r_lib/')
dir.create(temp_lib_path)
.libPaths(temp_lib_path)
install.packages(c('remotes', 'reshape2'))
remotes::install_github("lionel-/ggvis@fix-dplyr-1-0-0")

#Load the ggplot2 package
library(reshape2)
library(ggvis)
```


## Acquire dataset

We want to start by loading our dataset. We will use the 2nd Quarter 2020 (Advance Estimate) and Annual Update release as downloaded from the BEA site.

If we look at the data, we notice that we will need to skip some rows--4 rows--at the beginning of the file, or else we will get an error, so we use the "skip" option.

We also observe that loading it with headers creates variable names out of the years, and appends name with number as "[year].[#]" if there is more than one column for the year. We can use this to rename our variables.

```{r gdp}
data <- read.csv('data/national_gdp.csv', skip=4)
#Now look at the data
data

# Use the firstrow to rename the columns
new_names_year <- unlist( #unlist unravels list structure with depth 1
  lapply( #lapply applies a function to a list
    strsplit(names(data), split='.', fixed=T), #strsplit will return a list of vectors split from input vector on "." 
    function(x){ #this is the function being applied to strsplit return list by lapply
      return(x[1]) #return the first item ofthe vector--the year, preceded by "X" (or "Line"/"X" for first 2 columns)
  })
)

#Join new names with the first row 
new_names_qtrs <- paste0(new_names_year, as.character(unlist(data[1,])))

names(data) <- new_names_qtrs

#delete the first row!
gdp_raw <- data[-c(1), ]
gdp_raw
```

## Just for fun, let's also get quarterly population estimates.
```{r popdata}
pop <- read.csv('data/national_population.csv')

pop$Period <- paste0(substr(pop$DATE, 1, 4), 'Q', round(as.numeric(substr(pop$DATE, 6, 7))/3)+1)

pop

```

## Reshape the data for visualization

We have already made one parsing decision that has reshaped our data by omission with our skip option.

Now we are going to have to change it into a "clean" format: Every column is a variable name, every row is an observation.

We can do this more easily with the "reshape2" library.

```{r reshape}
# Melt to long format with line number and component as ID variables
gdp_long <- melt(gdp_raw, id.vars=c('LineLine', 'X'))

# Now is a good time to rename our variables
names(gdp_long) <- c('Line', 'Component', 'Period', 'Dollars')

# Remove the "X" from the period
gdp_long$Period <- gsub('X', '', gdp_long$Period, fixed = T)

# For now, we really only want top line GDP--line 1.
# We make a new data.frame from the old, ensuring our data types are correct.

# Also, we can add our pop data in now; I have been kind and set it up so that the dates are the same, but usually you would want to use a join to make sure the dates match for each row.

gdp <- data.frame(
  Period = as.character(gdp_long[gdp_long$Line==1,]$Period), 
  Dollars = as.numeric(gdp_long[gdp_long$Line==1,]$Dollars),
  Population = pop$Population
)

gdp

```

```{r ggvis_slide_demo}

getTooltipVals <- function(d){
   #Uses HTML
  return(paste0('Species: ', d$Species, '<br>Sepal Length: ', d$Sepal.Length, '<br>Sepal Width: ', d$Sepal.Width))
}

p_iris <- iris %>% 
  ggvis(x = ~Sepal.Length, y = ~Sepal.Width, fill = ~Species) %>%
	layer_points() %>% 
	scale_numeric("x") %>% 
	scale_numeric("y")%>% 
  add_tooltip(function(d){
    return(getTooltipVals(d))
    }) #Use HTML

p_iris_smooth <- iris %>% 
  ggvis(x = ~Sepal.Length, y = ~Sepal.Width) %>%
	layer_points() %>% 
  layer_smooths(span = 0.3, stroke := "red") %>%
	layer_model_predictions(model = "lm", se = TRUE) %>% 
	scale_numeric("x", trans = "log") %>% 
	scale_numeric("y", trans = "log") %>% 
  add_tooltip(getTooltipVals)

p_iris
p_iris_smooth

```

```{r modifying_ggplot}
annotation <- data.frame(
   Period = c("2007Q4","2009Q2", "2020Q2"),
   Dollars = c(20000,10000, 25000),
   label = c("Start of G.R.", "End of G.R.", "COVID Recession")
)


add_title <- function(vis, ..., x_lab = "X units", title = "Plot Title") 
{
  add_axis(vis, "x", title = x_lab) %>% 
    add_axis("x", orient = "top", ticks = 0, title = title,
             properties = axis_props(
               axis = list(stroke = "white"),
               labels = list(fontSize = 0)
             ), ...)
}

p <- ggvis(gdp, x=~Period, y=~Dollars) %>% # our data and mapping layer
#  annotate("rect", xmin = '2007Q4', xmax = '2009Q2', ymin = 0, ymax = 25000, color = "orange", fill = "orange", size = 0.5, alpha=0.4) + # Our annotation layers go first!
#  annotate("pointrange", x = '2020Q2', y = 20000, xmin = '2020Q1', xmax = '2020Q2', ymin = 0, ymax = 25000, color = "red", size = 1.5, alpha=.75) +
  #layer_lines(size = ~Dollars/Population, stroke = "#FC4E07", alpha = 0.5) %>%  # Our two geom line (mark) layers
  layer_lines() %>%
#  layer_smooths(span = 1, stroke := "red") %>% # Our stats layer; ggvis actually defaults to LOESS
#	layer_model_predictions(model = "loess", se = TRUE) %>% 
  scale_numeric("y") %>% #, labels=scales::dollar_format()) %>%  # Use the "dollar_format()" function from scales (ggplot dep'cy)
  scale_ordinal("x") %>% #, breaks = dates_to_show) %>% # Use the dates_to_show var we defined above to skip 20 quarters
  layer_text(data=annotation, x=~Period, y=~Dollars, baseline=~label,   #Add our text annotations AFTER all other
            fontSize=3 , angle=45) %>% #, stroke="black",                          #  geoms so it is on top of all other marks
#            fontface="bold", hjust = 1) + 
  add_title(title = "Nominal Gross Domestic Product \nTotal [Billions], Per Capita, and Projection")         #Add chart title, subtitle  
#  theme(axis.text.x = element_text(angle = 45, hjust=1), #Rotate axis labels and shift to the right
#        plot.title = element_text(hjust = 0.5),          #Center chart title
#        plot.subtitle = element_text(hjust = 0.5))       #Center chart subtitle

p

```