---
title: 'Visualization in R Exercise 2: ggvis'
author: "Andrea Batch"
date: "August 18, 2020"
output: html_document
---

```{r workdir, include=FALSE}

#You must change this!
WORKING_DIRECTORY <- 'CHANGE THIS TO YOUR WORKING DIRECTORY (top level folder where you cloned git repo)'

#Not changed
.libPaths('\\\\serv03cl\\data\\APPS\\R\\libraries_3_6_3_C')

```

Setup some requirements
```{r setup, include=FALSE}
#Set working directory (also do this for knit)
setwd(WORKING_DIRECTORY)
knitr::opts_knit$set(root.dir = WORKING_DIRECTORY)
knitr::opts_chunk$set(echo = TRUE)

#remotes::install_github("lionel-/ggvis@fix-dplyr-1-0-0") #Not run

#Load the ggplot2 package
library(reshape2)
library(ggvis)
```


## Acquire dataset

We want to start by loading our dataset. We will use the 2nd Quarter 2020 (Advance Estimate) and Annual Update release as downloaded from the BEA site.

If we look at the data, we notice that we will need to skip some rows--4 rows--at the beginning of the file, or else we will get an error, so we use the "skip" option.

We also observe that loading it with headers creates variable names out of the years, and appends name with number as "[year].[#]" if there is more than one column for the year. We can use this to rename our variables.

```{r gdp}
data <- read.csv('data/national_gdp.csv', skip=4)
#Now look at the data
data

# Use the firstrow to rename the columns
new_names_year <- unlist( #unlist unravels list structure with depth 1
  lapply( #lapply applies a function to a list
    strsplit(names(data), split='.', fixed=T), #strsplit will return a list of vectors split from input vector on "." 
    function(x){ #this is the function being applied to strsplit return list by lapply
      return(x[1]) #return the first item ofthe vector--the year, preceded by "X" (or "Line"/"X" for first 2 columns)
  })
)

#Join new names with the first row 
new_names_qtrs <- paste0(new_names_year, as.character(unlist(data[1,])))

names(data) <- new_names_qtrs

#delete the first row!
gdp_raw <- data[-c(1), ]
gdp_raw
```

## Just for fun, let's also get quarterly population estimates.
```{r popdata}
pop <- read.csv('data/national_population.csv')

pop$Period <- paste0(substr(pop$DATE, 1, 4), 'Q', round(as.numeric(substr(pop$DATE, 6, 7))/3)+1)

pop

```

## Reshape the data for visualization

We have already made one parsing decision that has reshaped our data by omission with our skip option.

Now we are going to have to change it into a "clean" format: Every column is a variable name, every row is an observation.

We can do this more easily with the "reshape2" library.

```{r reshape}
# Melt to long format with line number and component as ID variables
gdp_long <- melt(gdp_raw, id.vars=c('LineLine', 'X'))

# Now is a good time to rename our variables
names(gdp_long) <- c('Line', 'Component', 'Period', 'Dollars')

# Remove the "X" from the period
gdp_long$Period <- gsub('X', '', gdp_long$Period, fixed = T)

# For now, we really only want top line GDP--line 1.
# We make a new data.frame from the old, ensuring our data types are correct.

# Also, we can add our pop data in now; I have been kind and set it up so that the dates are the same, but usually you would want to use a join to make sure the dates match for each row.

gdp <- data.frame(
  Period = as.character(gdp_long[gdp_long$Line==1,]$Period), 
  Dollars = as.numeric(gdp_long[gdp_long$Line==1,]$Dollars),
  Population = pop$Population
)

gdp

```

```{r ggvis_slide_demo}
slider <- input_slider(10, 1000)

getTooltipVals <- function(d){
  message(paste(names(d), collapse = ', ')) #message--showing you the columns of d 
  message(paste(d, collapse = ', ')) #message--showing you the values of d 

  #Uses HTML for rendering tooltip
  return(paste0('Species: ', d$Species, '<br>Sepal Length: ', d$Sepal.Length, '<br>Sepal Width: ', d$Sepal.Width))
}

p_iris <- iris %>% 
  ggvis(x = ~Sepal.Length, y = ~Sepal.Width, fill = ~Species) %>%
	layer_points(size:=slider) %>% 
	scale_numeric("x") %>% 
	scale_numeric("y")%>% 
  add_tooltip(function(d){
    return(getTooltipVals(d))
    }) #Use HTML

p_iris
```

```{r ggvis_slide_demo2}
p_iris_smooth <- iris %>% 
  ggvis(x = ~Sepal.Length, y = ~Sepal.Width) %>%
	layer_points() %>% 
  layer_smooths(span = 0.3, stroke := "red") %>%
	layer_model_predictions(model = "lm", se = TRUE) %>% 
	scale_numeric("x", trans = "log") %>% 
	scale_numeric("y", trans = "log") %>% 
  add_tooltip(getTooltipVals)

p_iris_smooth

```

```{r modifying_ggplot}

#Two options here: We either change the date format to months and read as.Date, or we use the dates as ordinal strings; the latter does not work as well with ggvis, but I will use it just to demonstrate why. 
annotation <- data.frame(
   PeriodStart = c("2007Q4","2020Q1"),
   PeriodEnd = c("2009Q2","2020Q2"),
   DollarsMin = c(0,0),
   DollarsMax = c(max(gdp$Dollars),max(gdp$Dollars)),
   label = c("Great Recession", "COVID Recession")
)

add_title <- function(vis, ..., x_lab = "X units", title = "Plot Title") 
{
  add_axis(vis, "x", title = x_lab) %>% 
    add_axis("x", orient = "top", ticks = 0, title = title,
             properties = axis_props(
               axis = list(stroke = "white"),
               labels = list(fontSize = 0)
             ), ...)
}


#We will filter the data using this subset of rows
rowsToKeep_no_annot <- (1:dim(gdp))[1:(dim(gdp)[1]) %% 12 == 0]

#But we also want to make sure we have our annotation dates; we will add the index for each, and then make sure we have a unique set of row numbers, and then sort.
rowsToKeep <- c(
    rowsToKeep_no_annot, #Existing rows to keep
    (1:dim(gdp)[1])[     #vector of row numbers
      gdp$Period %in% c(as.character(annotation$PeriodStart), as.character(annotation$PeriodEnd)) #check if period in our annotations table start and end dates
      ] 
    ) %>% unique %>% sort #ensure it is unique and sorted

#And we want to use our annotations with a tooltip this time
getAnnotTooltips <- function(d){
  message(paste(names(d), collapse = ', ')) #message--showing you the columns of d 
  message(paste(d, collapse = ', ')) #message--showing you the values of d 

  #Uses HTML for rendering tooltip
  return(paste0('<b>', d$label, '</b><br>(', d$PeriodStart, ' - ', d$PeriodEnd))
}

p <- ggvis(data=gdp[rowsToKeep,], x=~Period, y=~Dollars) %>% # our data and mapping layer
  layer_paths() %>%
  add_axis(
    'x', 
    properties = axis_props(
      labels = list(angle=90, fontSize = 10)
  )) %>%
  add_axis('y', title='Nominal Dollars') %>%
  scale_ordinal("x", domain=as.character(gdp$Period)[rowsToKeep]) %>%
    layer_rects(data=annotation, x=~PeriodStart, x2=~PeriodEnd, y=~DollarsMin, y2=~DollarsMax, fillOpacity:=0.5, fill:="orange", stroke:=NA, opacity:=~label) %>% #We include opacity as label just to keep our label var for tooltip
  add_tooltip(function(d){
    return(getAnnotTooltips(d))
  })  %>%
  add_title(title = "Nominal Gross Domestic Product \nTotal [Billions], Per Capita, and Projection")         #Add chart title, subtitle  

p

```